Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-19 thesis/01-intro.Rmd
Adds remarks for chapter 2

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of data to learn from. This data originates from human activities and historical recordings such as texts, messages or speeches. It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models. There are many different types of models in NLP depending on a task that they are supposed to solve. However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models. The models usually assign the values based on the context in which the words appear. This means that the input data can have enormous influence on the outcome. The biases seem to have their primary source in the way the words are assigned numerical values.

>>>>

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of data to learn from. This data originates from human activities and historical recordings such as texts, messages or speeches. It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. There are many different types of models in NLP depending on a task that they are supposed to solve. However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings. The word embeddings are usually assigned the values based on the context in which the words appear. This means that the input data can have an enormous influence on the outcome. The biases seem to have their primary source in the way the words are assigned numerical values. One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by for example modifying the word embeddings directly.

### 2021-06-19 thesis/02-cosine.Rmd
Adds remarks for chapter 2

To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form. In the field of natural language processsing there are two main types of words representation --- localist and distributed. One-hot encoding is an example of a method used to achieve a localist representation of words. Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1. An example of a localist representation is:

@@

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word the model tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

@@

It is worth mentioning one more point concerning cosine similarity. After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
<!-- add information about inner product role -->
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.

@@

The greater the similarity between two vectors, the smaller the distance between them. The cosine distance ranges between 0 and 2. If the vectors are in an opposite direction to each other, the cosine distance is 2. And if the vectors are extremely similar then the cosine distance is very close to 0.

One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in [Statexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words. The plot below does not originate from the original paper (it is from https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings) but similar visualization may be found there. Data used to create our plot is as follows.

@@

The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he). Inner product of two vectors expresses similarity between words. This assumption originates from the geometry and properties of a vector space.

@@

We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.

@@

In the article there was no analysis of individual distances but the general look at the data through the usage of mean. The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race. In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam). Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.

In the article the authors also try to remove previously defined biases from word embedding. First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction. It is applied here to choose the subspace that contains the greatest amount of information. There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:

@@

The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words. This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed. In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.

After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed. In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.

In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning. If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved. As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.

@@

There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset. There are however some ways to estimate how many samples we need to make sure that the result is significant. Our research will show that the numbers used
are rather insufficient.

\textbf{2. No control while taking the mean of cosines}

The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. 
They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.

@@

A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm. Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

@@

In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. @Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion). We would like to focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

@@

MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.

@@

In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
[Analyticsindiamag Curse of dimensionality](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms. One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data. At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.

>>>>

To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form. In the field of natural language processsing there are two main types of words representation --- localist and distributed. One-hot encoding is an example of a method used to achieve a localist representation of words. Here, each vector contains information only about a single data point. This is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1. An example of a localist representation is:

@@

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word the model tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

@@

It is worth mentioning one more point concerning cosine similarity. After the vectors are normalized to have length equal to 1, the inner product itself (often dot product) is used to measure the similarity, because it then equals the cosine similarity.

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
<!-- add information about inner product role -->
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity. It is also possible to obtain negative cosine similarity, if the value approaches -1 it means that the words are contrary to each other.

@@

The greater the similarity between two vectors, the smaller the distance between them. The cosine distance ranges between 0 and 2. If the vectors are in an opposite directions, the cosine distance is 2. And if the vectors are extremely similar then the cosine distance is very close to 0.

One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words. The plot below does not originate from the original paper but from [kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings). However, similar visualization may be found there. Data used to create our plot is as follows.

@@

The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he). Inner product of two vectors expresses similarity between words. This assumption originates from the geometry and properties of a vector space. In the picture one may observe a correlation between the occupation and gender. Stereotypical male professions are closer to the "he" word and stereotypical female professions have greater similarity with "she" word. One could argue that this is a hint that there is some hidden bias information stored in the words vectors.

@@

We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown in Table 2.3

@@

In the article there was no analysis of individual distances, but the general look at the means. The authors introduced a metric that tries to generalize WEAT in measuring the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race. In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam). Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.

In the article the authors also try to remove biases, as previously defined, from a word embedding. First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction. It is applied here to choose the subspace that contains the greatest amount of information. There can be many subspaces found in a given group. For example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:

@@

The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words. This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed. In some cases this knowledge is useful, but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.

After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed. In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and stereotype is smaller.

In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning. If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved. As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.

@@

There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset. There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present. Our research will show that the numbers used
are rather insufficient.

\textbf{2. No attention to distributions and details}

The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes. They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class. If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words. We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we get the proper insight into the data.

@@

With such a method the uncertainty is not included which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

@@

In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. @Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion). For now let us focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

@@

MAC stand for mean average cosine similarity, although in reality the the table contains mean distances. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. In contrast, proper intervals decide whether given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals.

@@

In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
[Curse of dimensionality at analyticsindiamag.com](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms. One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data. At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
